Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-24 paper/paper.tex
standardize quotes

\section{Introduction}
%
In the absence of external stimulation, the human brain is not at rest.
In the beginning of the 21st century,
brain imaging may have been the first technique
to allow for the discovery of a unique brain network that probably maintains
baseline mental activities
\citep{raichle2001pnas, randy2008, bzdok2015resting}.
The "default mode network" (DMN) continues
to metabolize large quantities of
oxygen and glucose energy to maintain
neuronal computation
in the absence of behavioral focus
\citep{kenet2003spontaneously, fiser2004small}.
The baseline energy demand is only weakly modulated
at the onset of defined psychological tasks \citep{raichle_baseline}.
At the opposite, during sleep, the
decoupling of brain structures discarded the idea of the DMN being
only a passive network resonance and rather supported an
important role in sustaining conscious awareness \citep{horovitz2009decoupling}.
This dark matter of brain physiology begs the question
of the biological purpose underlying DMN activity
\citep{raichle2006brain}.
What has early been described as the "stream of consciousness"
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
This set of some of the most advanced regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
may exert higher-order control on human behavior.
It follows the notion of "a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex" \citep{carhart2010default}.

@@

The PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication \citep{flechsig1920}.
Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
%
Indeed,
the PMC has long been speculated to reflect constant computation of
environmental statistics and its internal representation
as an inner "mind's eye" \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological significance assessment.

@@

The TPJ in the left-hemispheric hemisphere (LTPJ),
in turn, exhibits a close relationship to Wernicke's area
involved in understanding semantics, such as in spoken and written language.
Neurological patients with damage in Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Patient speech
preserves natural rhythm and normal syntax, yet the
voiced sentences lack meaning (i.e., aphasia).
Abstracting from the typical speech interpretations in linguistics
and neuropsychology,
the LTPJ probably mediates access to and integration of world knowlege,
such as required during action considerations
\citep{binder2011neurobiology, seghier2013angular}.
For instance, LTPJ damage in humans also entail problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing explicit language content
\citep{varney1987locus}.
% (Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Inner speech also hinges on knowledge recall
about the physical and social world.
Indeed,
the internal generation of
verbalized thought ("language of the mind") was closely related to the LTPJ
in a pattern analysis of brain volume
\citep{geva2011neural}.
Further,
episodic memory recall and imagination strongly draw on
reassembling world knowledge.
Isolated building blocks of world statistics probably get reassembled
in internally generated mental scenarios that
guide present action choice,
weigh hypothetical possibilities, and forecast the future.
%
The LTPJ may hence facilitate the automated prediction of events
by incorporating experience-derived building blocks of world statistics
into ongoing action, planning, and problem solving.
% world -> policy matrix driven computation
% self -> value matrix driven computation

@@

\subsection{Markov Decision Processes}
We argue that the brain in general and the DMN in particular
are a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state but also on some finite
number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

@@

\subsubsection{The objective value of an action given a state}
Given only the limited information available from a POMDP, at a state $s$ the average quality or usefulness of taking an action $a$ under a policy $\pi$ is completely captured by the single number
% We can now formulate the purpose of the DMN
% as finding a policy $\pi$ for the human agent that maximizes the
% expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a],
\end{equation}
called the $Q$-value for the state-action pair $(s,a)$.
In other words, $Q^{\pi}(s,a)$ corresponds to the expected reward
over all considered action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$, and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utility of taking action $a$ in this situation ?".
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior that
may well reflect processing aspects of the DMN in human agents.

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an oracle which maps from states directly to actions. Here $\tilde{Q}(s,a|\theta)$ is an
approximate version of the $Q$-value function parameterized by the tensor $\theta$.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simplistic linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC (BA9) would implement such a greedy lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}.
  The oracle Q-values $\tilde{Q}(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state $s$.
  The DMN, and in particular its dmPFC node, would then contribute to the choice, at a given state $s$, of an action $a$ which maximizes this tabulated values. This mapping from states to actions is conventionally called "policy matrix". Learning consists in starting from a given instantiation of such a table and tweaking it around while using it to take actions which take the agent to different areas of the table, and so forth.

>>>>

\section{Introduction}
%
In the absence of external stimulation, the human brain is not at rest.
In the beginning of the 21st century,
brain imaging may have been the first technique
to allow for the discovery of a unique brain network that probably maintains
baseline mental activities
\citep{raichle2001pnas, randy2008, bzdok2015resting}.
The ``default mode network'' (DMN) continues
to metabolize large quantities of
oxygen and glucose energy to maintain
neuronal computation
during task-unconstrained behavior
\citep{kenet2003spontaneously, fiser2004small}.
The baseline energy demand is only weakly modulated
at the onset of defined psychological tasks \citep{raichle_baseline}.
At the opposite, during sleep, the
decoupling of brain structures discarded the idea of the DMN being
only a passive network resonance and rather supported an
important role in sustaining conscious awareness \citep{horovitz2009decoupling}.

This dark matter of brain physiology begs the question
of the biological purpose underlying DMN activity
\citep{raichle2006brain}.
What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
We propose that this set of some of the most advanced regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control on human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex'' \citep{carhart2010default}.

@@

The PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication \citep{flechsig1920}.
Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
%
Indeed,
the PMC has long been speculated to reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological significance assessment.

@@

The TPJ in the left-hemispheric hemisphere (LTPJ),
in turn, exhibits a close relationship to Wernicke's area
involved in understanding semantics, such as in spoken and written language.
Neurological patients with damage in Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Patient speech
preserves natural rhythm and normal syntax, yet the
voiced sentences lack meaning (i.e., aphasia).
Abstracting from the typical speech interpretations in linguistics
and neuropsychology,
the LTPJ probably mediates access to and integration of world knowlege,
such as required during action considerations
\citep{binder2011neurobiology, seghier2013angular}.
For instance, LTPJ damage in humans also entail problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing explicit language content
\citep{varney1987locus}.
% (Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Inner speech also hinges on knowledge recall
about the physical and social world.
Indeed,
the internal generation of
verbalized thought (``language of the mind'') was closely related to the LTPJ
in a pattern analysis of brain volume
\citep{geva2011neural}.
Further,
episodic memory recall and imagination strongly draw on
reassembling world knowledge.
Isolated building blocks of world statistics probably get reassembled
in internally generated mental scenarios that
guide present action choice,
weigh hypothetical possibilities, and forecast the future.
%
The LTPJ may hence facilitate the automated prediction of events
by incorporating experience-derived building blocks of world statistics
into ongoing action, planning, and problem solving.
% world -> policy matrix driven computation
% self -> value matrix driven computation

@@

\subsection{Markov decision processes}
We argue that the brain in general and the DMN in particular
are a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state but also on some finite
number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

@@

\subsubsection{The objective value of an action given a state}
Given only the limited information available from a POMDP, at a state $s$ the average quality or usefulness of taking an action $a$ under a policy $\pi$ is completely captured by the single number
% We can now formulate the purpose of the DMN
% as finding a policy $\pi$ for the human agent that maximizes the
% expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a],
\end{equation}
called the $Q$-value for the state-action pair $(s,a)$.
In other words, $Q^{\pi}(s,a)$ corresponds to the expected reward
over all considered action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$, and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
``What is the expected utility of taking action $a$ in this situation ?''.
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior that
may well reflect processing aspects of the DMN in human agents.

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an oracle which maps from states directly to actions. Here $\tilde{Q}(s,a|\theta)$ is an
approximate version of the $Q$-value function parameterized by the tensor $\theta$.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simplistic linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC (BA9) would implement such a greedy lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}.
  The oracle Q-values $\tilde{Q}(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state $s$.
  The DMN, and in particular its dmPFC node, would then contribute to the choice, at a given state $s$, of an action $a$ which maximizes this tabulated values. This mapping from states to actions is conventionally called ``policy matrix''. Learning consists in starting from a given instantiation of such a table and tweaking it around while using it to take actions which take the agent to different areas of the table, and so forth.
